---
title: "Exploration of the Exponential Distribution"
author: "Sharon Francisco"
date: "July 20, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

This report shows how the distribution of averages of an exponential distribution approximates a normal distribution as the sample size increases because of the Central Limit Theorem (CLT).

## Installing Packages and Loading Libraries 
The packages used in the analysis are installed below.

```{r installAndLoadPackages}
if(!require(ggplot2, quietly = TRUE)) install.packages("ggplot2")
library(ggplot2)
```


## Exponential Distribution

To explore the exponential distribution the assignment says to use lambda = 0.2 as the rate parameter.

The code below creates the exponential distribution with lambda = 0.2 and puts it in a data frame. I will use this later to create a plot of the distribution.

```{r createExponenDistrib}
lambda = 0.2
x <- seq(0, 40, length.out = 1000)
px <- dexp(x, rate = lambda)
dat <- data.frame(x = x, px)
```

The mean, mu, of an exponential distribution is 1/lambda, and the standard deviation, sigma, is also 1/lambda.

```{r calculateMuSd}
mu <- 1 / lambda
mu
```

The variance, according to Wikipedia, is 1/lambda^2, which is sigma^2.

```{r calculateVariance}
sigma <- 1 / lambda

xpDistrVar <- 1 / lambda^2
xpDistrVar
```

The code below plots the exponential distribution with lambda, the rate, = 0.2, which is the y-intercept. It also shows the theoretical mean, mu, and the variance.

```{r plotExponentialDistrib}
with(dat, plot(x, px))
par(mar = c(3, 4, 4, 0))
title(main = "Exponential Distribution, lambda = 0.2")
text(0, lambda, labels = "lambda", pos = 4)
abline(v = mu, col = "green")
muLabel <- paste("mu =", mu)
text(mu, 0, labels = muLabel, pos = 4, col = "green")
abline(v = xpDistrVar, col = "blue")
xpDistVarLabel <- paste("exponential variance =", xpDistrVar)
text(xpDistrVar, 0.025, labels = xpDistVarLabel, pos = 1, 
     col = "blue")

```

So for an exponential distribution with lambda = 0.2, the mean and standard deviation are `r mu`, and the variance is `r xpDistrVar`.


## Distribution of 1000 Simulations of the Means of 40 Exponential Distributions

Now I'll look at the distribution of the *averages* of 40 exponentials simulated 1000 times.

```{r AvgOf40Simulated1000x}
set.seed(77)
n <- 40
simulations <- 1000
cLim <- replicate(simulations, mean(rexp(n, lambda)))
```

I'll calculate the sample mean of this disribution.

```{r calculateSampleMean}
sampleMean <- mean(cLim)
sampleMean
```

The sample mean, `r sampleMean`, is very close to the theoretical mean, `r mu`.

Next I'll calculate the theoretical variance using the formula, sigma^2 / n, and then the sample variance using the **var** function in R.

```{r calculateVariance2}
theoreticalVarn <- sigma^2 / n
theoreticalVarn
varnSample <- var(cLim)
varnSample
```

The results of the two variance calculations are close. Because we ran 1000 simulations, the variance is relatively small. Because this is a distribution of averages, the variance is very different than the variance of the original exponential distribution.

I'll plot a histogram of this distribution of the *averages* of `r n` exponentials simulated 1000 times.

```{r histogramAvgOf40Simulated1000x}
hist(cLim)
abline(v = sampleMean, col = "red")
mnClimLabel <- paste("sample mean =", sampleMean)
text(sampleMean, 250, labels = mnClimLabel, pos = 1,
     col = "red")
abline(v = mu, col = "green")
text(mu, 250, labels = muLabel, pos = 4, col = "green")

```

Since this is a distribution of the averages of `r n` distributions, the distribution looks Gaussian because of the Central Limit Theorem, which states that the distribution of averages of iid variables becomes that of a standard normal as the sample size increases.

Here are the density plots of the distribution of averages using the sample mean (`r sampleMean`) and sample variance (`r varnSample`) and the theoretical mean, mu (`r mu`), and the theoretical variance (`r theoreticalVarn`). 

```{r plotVariances}
g <- ggplot(as.data.frame(cLim), aes(cLim)) +
        xlab("Means") +
        ylab("Density")
g <- g + stat_function(fun = dnorm,
                       args = list(mean = sampleMean, 
                                   sd = sqrt(varnSample)),
                       aes(color = "Sample Variance"), 
                       size = 1) +
        stat_function(fun = dnorm,
                      args = list(mean = mu, 
                                  sd = sqrt(theoreticalVarn)),
                      aes(color = "Theoretical Variance"), 
                       size = 1)
g <- g + geom_vline(aes(xintercept = sampleMean, color = 
                                "Mean of Sample Means")) +
        geom_vline(aes(xintercept = mu, color = 
                               "Mu (Theoretical Mean)"))
g

```

This demonstrates the characteristics of the Central Limit Theorem:

1. The mean of the sample means is approximately equal to the theoretical mean;

2. The sample variance is approximately equal to the theoretical variance divided by population size;

3. The distribution of the sample means is approximately normal.


























